Communication complexity: Chapter 3 \\ Distributional communication complexity

### 2019-04-13 CS860-CommunicationComplexity/assignments/Chapter3.tex
Corruption bound

Consider a uniformly random $r \in \{0,1\}$. If $x=y$ then $\pi_r(x,y) = 1$. If $x\neq y$ then $\pi_r(x,y) =1$ with probability $\frac{1}{4}$. Let $X_r(x,y)$ be the indicator random variable indicating if $\pi_r(x,y)$ fails on input $(x,y)$. Thus we see that for any $(x,y)$,
$$\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
Thus we have
$$\mathbb{E}_{(x,y)\sim \mu}\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
Rearranning order of expectation yields
$$\mathbb{E}_{r}\mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
And thus there exists an $r \in \{0,1\}$ such that
$$\Pr_\mu(\pi_r \text{ fails}) \leq \mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
\end{proof}

@@

\begin{lemma}
Fix a function $f : \calX \times \calY \to \{0,1\}$ and a distribution
$\mu$ on $\calX \times \calY$.
If there exist parameters $\alpha, \beta > 0$ for which
every rectangle $R \subseteq \calX \times \calY$ satisfies 
\[
\mu(R \cap f^{-1}(0)) \ge \alpha \cdot \mu(R) - \beta
\]
then
\[
D^\mu_\epsilon(f) \ge \log\left( \frac{\alpha \cdot \mu(f^{-1}(1)) - \epsilon}{\beta}\right).
\]
\end{lemma}

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$.
Let $\{R_1, \dots, R_t\}$ be set of nearly $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf.

$$1= \sum_{i=1}^t \mu(R_i) = \sum_{i=1}^t \mu(R_i \cap f^{-1}(0)) + \mu(R_i\cap f^{-1}(1)) \geq \alpha - t \beta +\mu(f^{-1}(1))$$
Thus
$$t \geq \frac{\alpha + \mu(f^{-1}(1)) -1}{\beta}.$$
Now we claim that 
$$\alpha + \mu(f^{-1}(1)) -1 \geq \alpha\cdot \mu(f^{-1}(1)) -\epsilon.$$
This claim is equivalent to
$$1-\epsilon \leq \alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)).$$
Since 
$$\alpha(\mu(f^{-1}(0))) +\mu(f^{-1}(1)) = 1-(1-\alpha) \mu(f^{-1}(0))$$
the claim holds since the probability of failure is at least $(1-\alpha)f^{-1}(0)$ \replacethistext{(I didn't have time to formalize)}.
\end{proof}

>>>>

Consider a uniformly random $r \in \{0,1\}$. If $x=y$ then $\pi_r(x,y) = 1$. If $x\neq y$ then $\pi_r(x,y) =1$ with probability $\frac{1}{4}$. Let $X_r(x,y)$ be the indicator random variable indicating if $\pi_r(x,y)$ fails on input $(x,y)$. Thus we see that for any $(x,y)$,
$$\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
Thus we have
$$\mathbb{E}_{(x,y)\sim \mu}\mathbb{E}_r[X_r(x,y)] \leq \frac{1}{4}.$$
Rearranging order of expectation yields
$$\mathbb{E}_{r}\mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
And thus there exists an $r \in \{0,1\}$ such that
$$\Pr_\mu(\pi_r \text{ fails}) \leq \mathbb{E}_{(x,y)\sim \mu}[X_r(x,y)] \leq \frac{1}{4}.$$
\end{proof}

@@

\begin{lemma}
Fix a function $f : \calX \times \calY \to \{0,1\}$ and a distribution
$\mu$ on $\calX \times \calY$.
If there exist parameters $\alpha, \beta \in (0,1)$ for which
every rectangle $R \subseteq \calX \times \calY$ satisfies 
\[
\mu(R \cap f^{-1}(0)) \ge \alpha \cdot \mu(R) - \beta
\]
then
\[
D^\mu_\epsilon(f) \ge \log\left( \frac{\alpha \cdot \mu(f^{-1}(1)) - \epsilon}{\beta}\right).
\]
\end{lemma}

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$.
Let $\mathcal{R}$ be the set of rectangles corresponding to the path from the root of $\pi$ to each respective leaf. For $b \in \{0,1\}$ let $\mathcal{R}_b \subseteq \mathcal{R}$ be the set of rectangles such that the protocol outputs $b$ at the corresponding leaf.

Since $\pi$ has success probability at least $1-\epsilon$,
$$\sum_{R \in \mathcal{R}_0} \mu(R\cap f^{-1}(1)) + \sum_{R \in \mathcal{R}_1} \mu(R\cap f^{-1}(0)) \leq \epsilon.$$
By the lemma hypothesis,
$$\sum_{R \in \mathcal{R}_1}\mu(R\cap f^{-1}(0)) \geq \alpha\sum_{R \in \mathcal{R}_1}\mu(R) -\beta\left|\mathcal{R}_1\right|.$$
Rearranging and combining the two inequalities above yields
\begin{align*}
    \beta|\mathcal{R}_1| &\geq \alpha\sum_{R \in \mathcal{R}_1}\mu(R) - \epsilon + \sum_{R \in \mathcal{R}_0}\mu(R\cap f^{-1}(1))\\
    &=\alpha \mu(f^{-1}(1)) - \epsilon + (1-\alpha)\sum_{R \in \mathcal{R}_0}\mu(R\cap f^{-1}(1))+\alpha \sum_{R \in \mathcal{R}_1}\mu(R\cap f^{-1}(0))\\
    &\geq \alpha \mu(f^{-1}(1)) - \epsilon
\end{align*}
The last inequality above follows since $\alpha, 1-\alpha$ are non-negative and $\mu$ is a probability distribution. Since $|\mathcal{R}_1| \leq |\mathcal{R}| = 2^{D^\mu_\epsilon(f)}$ we have the desired result.
\end{proof}
